On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-12-02 paper.tex
Rewriting

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with uniformly random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively. The orientation $\theta$ of the segment is uniformly distributed on $[0,2\pi)$. The variables $x_1$, $y_1$ and $\theta$ determine the coordinates $x_2$, $y_2$ of the second endpoint.

Each realization of the random segment gives rise to a discrete bounding rectangle, whose normalized dimensions $i$ and $j$ are thus random variables, as is the number $\tiles$ of visited tiles. Except for a set of realizations with zero probability, $i$ and $j$ are at least $1$, and $\tiles = i+j-1$ tiles. Note that $i$ and $j$ are not statistically independent.

Rectangular grids with real-valued segment lengths are considered in \S\ref{part: probmax: arbitrary grid, real lengths}. The main result is the average number of visited tiles. The square grid with real-valued lengths is addressed in \S\ref{part: probmax: unit square grid, real lengths}. This more specialized setting allows computation of the probability of visiting the maximum possible number of tiles.

@@

Let $\funta: \mathbb R^+ \to \mathbb R^+$ be defined such that $\funta(\len)$ gives the average number of tiles visited by a random segment of length $\len$, with the distributions specified in \S\ref{part: rand}. This function can be easily computed using the fact that
\begin{equation}
\label{eq: funta E[i] E[j]}
\funta(\len) = \E[\tiles] = \E[i]+\E[j]-1,
\end{equation}
and noting that events of probability $0$, such as $x_1=a$ or the segment passing through a grid point, can be disregarded.

\begin{theorem}
\label{theo: funta, form}
Given $a, b, \len \in \mathbb R^+$, consider a grid with parameters $a, b$ and a uniformly random segment of length $\len$, as defined above. The average number of tiles visited by the segment, $\funta(\len)$, is 
\begin{equation}
\label{eq: funta, form}
\funta(\len) = \frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right) + 1.
\end{equation}
\end{theorem}

@@

Consider again a grid with horizontal spacing $a$ and vertical spacing $b$. The grid crossings decompose into crossings of horizontal and vertical grid lines. By linearity of expectation, the expected number of crossings $\funca(\len)$ is the sum of the expectations for parallel lines with spacing $a$ and $b$ respectively, which gives
\begin{equation}
\label{eq: funca}
\funca(\len) = \frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right).
\end{equation}
As noted in Proposition \ref{prop: i+j-1}, the number of tiles visited by a segment is the count of its grid line crossings plus $1$, unless it exactly passes through any grid points, but that occurs with probability $0$ and therefore does not affect the expectation. Thus, $\funta(\len)$ is given by the right-hand side of \eqref{eq: funca} plus $1$.
\end{proof}

@@

The results in Theorem~\ref{theo: funta, form} and Proposition~\ref{prop: Pr i} make clear the relationship between the problem considered in this section here and Buffon's needle problem, i.e.~the probability that a random segment of fixed length crosses a line in a regular structure of parallel strips \cite[section 1.1]{Mathai99}. Firstly, since the number $\pi$ is involved in \eqref{eq: funta, form}, it is possible to design a simple probabilistic experiment to estimate the value of $\pi$, as in Buffon's original experiment. For example, choosing $a=b=\len=1$ gives $\funta(\len) = 1+1/\pi$.

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute, due to the irregularity of the optimal $(i,j)$ pairs (see \S\ref{part: fund results}). In the square case, however, the simplicity of the set makes the problem more tractable.

>>>>

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with uniformly random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively, where $a$, $b$ are the grid parameters. The orientation $\theta$ of the segment is uniformly distributed on $[0,2\pi)$. The coordinates $x_2$, $y_2$ of the second endpoint are determined by $x_1$, $y_1$ and $\theta$.

Each realization of the random segment gives rise to a discrete bounding rectangle, whose normalized dimensions $i$ and $j$ are thus random variables, as is the number $\tiles$ of visited tiles. Except for a set of realizations with zero probability, $i$ and $j$ are at least $1$, and $\tiles = i+j-1$. Note that $i$ and $j$ are not statistically independent.

This section deals with the two problems stated in \S\ref{part: intro} for random segments, namely obtaining the average number of visited tiles and the probability of visiting the maximum number of tiles. Segment lengths will be assumed to be real-valued. The results to be obtained are directly applicable for integer lengths as a particular case.

Arbitrary grids are considered in \S\ref{part: probmax: arbitrary grid, real lengths}. The main result is the average number of visited tiles. The square grid is addressed in \S\ref{part: probmax: unit square grid, real lengths}. This more specialized setting allows computation of the probability of visiting the maximum possible number of tiles.

@@

Let $\funta: \mathbb R^+ \to \mathbb R^+$ be defined such that $\funta(\len)$ is the average number of tiles visited by a random segment of length $\len$, with the distributions specified in \S\ref{part: rand}. This function can be easily computed using the fact that
\begin{equation}
\label{eq: funta E[i] E[j]}
\funta(\len) = \E[\tiles] = \E[i]+\E[j]-1,
\end{equation}
and noting that events of probability $0$, such as $x_1=0$ or the segment passing through a grid point, can be disregarded.

\begin{theorem}
\label{theo: funta, form}
Given $a, b, \len \in \mathbb R^+$, consider a grid with parameters $a, b$ and a uniformly random segment of length $\len$, as defined above. The average number of tiles visited by the segment is 
\begin{equation}
\label{eq: funta, form}
\funta(\len) = \frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right) + 1.
\end{equation}
\end{theorem}

@@

Consider again a grid with horizontal spacing $a$ and vertical spacing $b$. The grid crossings decompose into crossings of horizontal and vertical grid lines. By linearity of expectation, the expected number of crossings $\funca(\len)$ is the sum of the expectations for parallel lines with spacing $a$ and $b$ respectively, which gives
\begin{equation}
\label{eq: funca}
\funca(\len) = \frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right).
\end{equation}
As noted in Proposition \ref{prop: i+j-1}, the number of tiles visited by a segment is the count of its grid line crossings plus $1$, unless it exactly passes through any grid points, but that occurs with probability $0$ and therefore does not affect the expected value. Thus, $\funta(\len)$ is obtained by adding $1$ to \eqref{eq: funca}, which gives \eqref{eq: funta, form}.
\end{proof}

@@

The results in Theorem~\ref{theo: funta, form} and Proposition~\ref{prop: Pr i} make clear the relationship between the problem considered here and Buffon's needle problem, i.e.~the probability that a random segment of fixed length crosses a line in a regular structure of parallel strips \cite[section 1.1]{Mathai99}. Firstly, since the number $\pi$ is involved in \eqref{eq: funta, form}, it is possible to design a simple probabilistic experiment to estimate the value of $\pi$, as in Buffon's original experiment. For example, choosing $a=b=\len=1$ gives $\funta(\len) = 1+1/\pi$.

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute, due to the irregularity of the optimal $(i,j)$ pairs (as was discussed in \S\ref{part: fund results}; in particular see Figure~\ref{fig: pairs_bounds}). In the square case, however, the simplicity of the set of optimal pairs makes the problem more tractable.
